---
title: "Practical Machine Learning Assignment Writeup"
author: "Lars Skjærven"
date: "1/26/2018"
output: html_document
---

```{r setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

## Assignment

The goal of the project is to predict the exercise manner (or quality) of six participants based on a number of predictor variables. The outcome is stored in the `classe` variable of the training set. We will explore a several classifiers to solve this problem, notably **random forest**, **gradient boosting**, **linear discriminant analysis**, **support vector machine**, and **k-nearest neighbour**. Finally we will apply the final model to predict the 20 cases in the test set. 

In addition, we will calculate the importance of the variables based on our most successful model, and use the result of the importance to explore the underlying data. 


## Import data

Import required libraries, download data files, and use `read_csv()` to parse the csv files: 

```{r libs}
# Load required packages
library(caret)
library(readr)
library(corrplot)
```


```{r import, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
# Download data files
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", "pml-train.csv")
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", "pml-test.csv")

# Read data
train.orig <- read_csv("pml-train.csv")
train.orig$classe <- as.factor(train.orig$classe)
test.orig <- read_csv("pml-test.csv")
```


The training data consists of **`r dim(train.orig)[1]`** rows with **`r dim(train.orig)[2]`** columns. There are **`r dim(test.orig)[1]`** rows in the test set. The `classe` variable contains the following grades:

```{r overview}
# Table of outcome 'classe'
table(train.orig$classe)
```


## Data cleanup

The training dataset contains 7 columns that does not provide informative data for predicting the training outcome: **`r paste(colnames(train.orig)[1:7], sep=", ")`**. In addition, there are **`r sum(apply(train.orig, 2, function(x) !all(!is.na(x))))`** variables with a large number of missing (`NA`) values that we will ommit for further analysis:

```{r cleandata}
# Remove first 7 columns 
train <- train.orig
train <- train[, -c(1:7)]

# Remove NA-containing columns
col.inds <- apply(train, 2, function(x) all(!is.na(x)))
train <- train[, col.inds]
```

That leaves us with **`r ncol(train)-1`** variables in the training dataset. That is only **`r round( (ncol(train)-1) / (ncol(train.orig)-8), 2) * 100`** % of the variables of the original data set. To aovid this we could attempt to impute missing values, but we will check how our prediction will work without these variables before imputing. 


## Split data 

Let's split the training set into a test and training set using the `createDataPartition()` function of the R package:

```{r splitdata}
# Split data set 
inTrain <- createDataPartition(y=train$classe, p=0.7, list=FALSE)
mytest <- train[-inTrain,]
train <- train[inTrain,]
```

We now have **`r nrow(mytest)`** rows in the test set and **`r nrow(train)`** rows in the final training set. 

## Build models

Set up a common `trainControl` object to use in all models we explore. We will use a **5-fold cross-validation** in the following models. 

```{r traincontrol}
# set up a trainControl object with 5-fold cross-validation
trc <- trainControl(method="cv", number=5)
```


### Random forest

We start with exploring the **random forest** algorithm (provided with the **ranger** package):

```{r model_ranger}
# Ranom forrest
mod_rf <- train(classe ~ ., 
                method="ranger", data=train, 
                trControl=trc, 
                importance = 'impurity')

# Predict on test data
pred_rf <- predict(mod_rf, mytest)

# Cross-table
table(pred_rf, mytest$classe)

# Calcualte confusion matrix
cm_rf <- confusionMatrix(pred_rf, mytest$classe)
```

This model provides an accuracy of **`r round(cm_rf$overall["Accuracy"], 3)`**. That's pretty good... 


## Exploring important variables on prediction

So what's the important variables here? There are no clear correlations in the dataset:


```{r corrplot}
m <- cor(train[, -c(ncol(train))])
corrplot(m, method = "circle", tl.cex=0.5)
```


But we can use `varImp()` of the **caret** package to explore what's driving the prediction:

```{r}
# Find varialbe importance
rfImp <- varImp(mod_rf)
rfImp
```


Let's plot some of these important variables:

```{r}
# Plot important variables
ggplot(train, aes(classe, pitch_forearm)) +
  geom_boxplot(aes(col=classe)) 

ggplot(train, aes(classe, roll_belt)) +
  geom_boxplot(aes(col=classe)) 

ggplot(train, aes(classe, yaw_belt)) +
  geom_boxplot(aes(col=classe)) 

ggplot(train, aes(classe, roll_forearm)) +
  geom_boxplot(aes(col=classe)) 
```

It certainly looks like you need a `pitch_forearm` close to 0 for a `classe` A (median close to 0), `roll_belt` values of 0 (median close to 0), smaller `yaw_belt` (median close to -90), and a `roll_forearm` close to 0. 


## Final prediction on the test set

Based on the random forest model we predict the 20 data points of the test set:

```{r predict}
predict(mod_rf, test.orig)
```


## Exploring other models 

Although the random forest model provides an accuracy of **`r round(cm_rf$overall["Accuracy"], 3)`**, we will check if other popular algorithms performs equivalently good under default settings:  

### Gradient Boosting


```{r model_gbm}
# Boosting
mod_gbm <- train(classe ~ .,
                 method="gbm", data=train, 
                 trControl=trc, verbose=FALSE)

pred_gbm <- predict(mod_gbm, mytest)
table(pred_gbm, mytest$classe)
cm_gbm <- confusionMatrix(pred_gbm, mytest$classe)
cm_gbm$overall["Accuracy"]
```


### Linear Discriminant Analysis

```{r model_lda}
# Linear Discriminant Analysis
mod_lda <- train(classe ~ .,
                 method="lda", data=train, 
                 trControl=trc)

pred_lda <- predict(mod_lda, mytest)
table(pred_lda, mytest$classe)
cm_lda <- confusionMatrix(pred_lda, mytest$classe)
cm_lda$overall["Accuracy"]
```


### k-nearest neighbors

```{r knn}
mod_knn <- train(classe ~ ., 
                method = "knn",
                data = train, 
                trControl = trc)
pred_knn <- predict(mod_knn, mytest)
table(pred_knn, mytest$classe)
cm_knn <- confusionMatrix(pred_knn, mytest$classe)
cm_knn$overall["Accuracy"]
```

### SVM

```{r svm}
mod_svm <- train(classe ~ ., 
                method = "svmLinear",
                data = train, 
                trControl = trc)
pred_svm <- predict(mod_svm, mytest)
table(pred_svm, mytest$classe)
cm_svm <- confusionMatrix(pred_svm, mytest$classe)
cm_svm$overall["Accuracy"]
```


## Conclusion

Random forest seems to be an excellent model for predicting exercise manner with an accuracy of **`r round(cm_rf$overall["Accuracy"], 3)`**. k-NN gives an accuracy of **`r round(cm_knn$overall["Accuracy"], 3)`** while gradient boosting gives an accuracy of **`r round(cm_gbm$overall["Accuracy"], 3)`**. 


## Current session info

```{r}
# Information about the current session
sessionInfo()
```
