---
title: "Practical Machine Learning Assignment Writeup"
author: "Lars Skjærven"
date: "1/26/2018"
output: html_document
---

```{r setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

## Assignment

The goal of the project is to predict the exercise manner (or quality) of six participants based on a number of predictor variables. The outcome / manner is stored as the `classe` variable in the training set. We will explore a few classifiers to solve this problem, notably **random forest**, **gradient boosting**, **linear discriminant analysis**, **support vector machine**, and **k-nearest neighbour**. Finally we will apply the final model to predict the 20 cases in the test set. 


## Import data

Import required libraries, download data files, and use `read_csv()` to parse the csv files: 

```{r libs}
# Load required packages
library(caret)
library(readr)
library(corrplot)
```


```{r import, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
# Download data files
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", "pml-train.csv")
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", "pml-test.csv")

# Read data
train.orig <- read_csv("pml-train.csv")
train.orig$classe <- as.factor(train.orig$classe)
test.orig <- read_csv("pml-test.csv")
```


The training data consists of **`r dim(train.orig)[1]`** rows with **`r dim(train.orig)[2]`** columns. There are **`r dim(test.orig)[1]`** rows in the test set. 

```{r overview}
# Table of outcome 'classe'
table(train.orig$classe)
```


## Data cleanup

```{r cleandata}
# remove first 7 columns 
# id, user_name, timestamp, new_window
train <- train.orig
train <- train[, -c(1:7)]

# remove NA-containing columns
col.inds <- apply(train, 2, function(x) all(!is.na(x)))
train <- train[, col.inds]
```


## Split data 

to training and test set

```{r splitdata}
# split data set 
inTrain <- createDataPartition(y=train$classe, p=0.7, list=FALSE)
mytest <- train[-inTrain,]
train <- train[inTrain,]
dim(train)
dim(mytest)
```

## Explore data

```{r corrplot}
m <- cor(train[, -c(ncol(train))])
corrplot(m, method = "circle", tl.cex=0.5)
```



## Build models

```{r traincontrol}
# set up a trainControl object with 5-fold cross-validation
trc <- trainControl(method="cv", number=5)
```

### Random forest

```{r model_ranger}
# Ranom forrest
mod_rf <- train(classe ~ ., 
                method="ranger", data=train, 
                trControl=trc, 
                importance = 'impurity')

pred_rf <- predict(mod_rf, mytest)
table(pred_rf, mytest$classe)
cm_rf <- confusionMatrix(pred_rf, mytest$classe)
cm_rf$overall["Accuracy"]
```

### Boosting

```{r model_gbm}
# Boosting
mod_gbm <- train(classe ~ .,
                 method="gbm", data=train, 
                 trControl=trc, verbose=FALSE)

pred_gbm <- predict(mod_gbm, mytest)
table(pred_gbm, mytest$classe)
cm_gbm <- confusionMatrix(pred_gbm, mytest$classe)
cm_gbm$overall["Accuracy"]
```


### Linear Discriminant Analysis

```{r model_lda}
# Linear Discriminant Analysis
mod_lda <- train(classe ~ .,
                 method="lda", data=train, 
                 trControl=trc)

pred_lda <- predict(mod_lda, mytest)
table(pred_lda, mytest$classe)
cm_lda <- confusionMatrix(pred_lda, mytest$classe)
cm_lda$overall["Accuracy"]
```


### k-nearest neighbors

```{r knn}
mod_knn <- train(classe ~ ., 
                method = "knn",
                data = train, 
                trControl = trc)
pred_knn <- predict(mod_knn, mytest)
table(pred_knn, mytest$classe)
cm_knn <- confusionMatrix(pred_knn, mytest$classe)
cm_knn$overall["Accuracy"]
```

### SVM

```{r svm}
mod_svm <- train(classe ~ ., 
                method = "svmLinear",
                data = train, 
                trControl = trc)
pred_svm <- predict(mod_svm, mytest)
table(pred_svm, mytest$classe)
cm_svm <- confusionMatrix(pred_svm, mytest$classe)
cm_svm$overall["Accuracy"]
```



## Exploring important variables on prediction
```{r}
# Find varialbe importance
rfImp <- varImp(mod_rf)
rfImp
```


```{r}
# Plot important variables
ggplot(train, aes(classe, pitch_forearm)) +
  geom_boxplot(aes(col=classe)) 

ggplot(train, aes(classe, roll_belt)) +
  geom_boxplot(aes(col=classe)) 

ggplot(train, aes(classe, yaw_belt)) +
  geom_boxplot(aes(col=classe)) 

ggplot(train, aes(classe, roll_forearm)) +
  geom_boxplot(aes(col=classe)) 
```


## Predict on test set

```{r predict}
predict(mod_rf, test.orig)
```

